Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks.

### 2020-06-07 main.Rmd
added Grad-CAM part and remade some plots

---
title: "Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: svm_latex.tex
    toc: no
fontsize: 12pt
fontfamily: sans
papersize: a4paper
author:
  - name: Ruth Kristianingsih, Dan MacLean
  - affiliation: The Sainsbury Laboratory, University of East Anglia, Norwich Research Park, Norwich, UK, NR4 7JJ, dan.maclean@tsl.ac.uk
abstract: "TBD."
geometry: margin=1in
biblio-style: apsr
bibliography: "r-references.bib"
graphics: yes
header-includes:
  - \usepackage{amsmath}
---

@@

# Source plotting functions
source("functions/plot_confusion_mat.R")
source("functions/plot_confusion_matrices.R")
source("functions/get_lower_corr_mat.R")
source("functions/get_seqlogo.R")

@@

```{r analysis-preferences, echo=FALSE}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

@@

Sequence data were collected from the PHI-Base database version 4.8 [@phibase] by accessing a text dump of the data kindly prepared on request by the PHI-Base team, the file can be accessed at [https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv). The pipeline in Figure \@ref(fig:effectorcollection) outlines the steps used. We filtered plant effector proteins and their taxonomic groups and collected sequences from UniProt Release 2019_05, using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd). We created a correspondingly sized data set of non-effectors with secretion signals originating in species matched to those from which the effectors were drawn. We downloaded sequences for randomly selected proteins matching these criteria from Ensembl databases [@yates2020] : specifically Ensembl Fungi, Protists and Bacteria manually using the BioMart tools [@smedley2009]. Since the BioMart tool is not available on Ensembl Bacteria, we downloaded whole proteome protein sequnces from species matched to those from which the effector came using FTP. With these we used SignalP 3.0 [@dyrlovbendtsen2004] in order to filter the secreted sequences and selected accordingly. ^[Ruth, what version of the code and what parameter settings did you use?] Redundant sequences were filtered using BLASTp [@camachoblast2009]. We achieved these steps using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd).

```{r, effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt", message = FALSE, echo=FALSE}

knitr::include_graphics(
  "figures/flowchart_effectorsequence_collection.pdf",
)
```

```{r, noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists", message = FALSE, echo=FALSE}

knitr::include_graphics(
  "figures/flowchart_secreted_noneffectorsequence_collection.pdf",
)
```

@@

The sequences collected were encoded using either one-hot encoding (CNN-LSTM based models) or integer based encoding (CNN-GRU-LSTM models). Sequences were post-padded with zeroes to bring the vectors to identical lengths to each other and the longest sequence in the taxon data set ^[Ruth, what were the lengths for each model?]. Encoded sequences were split into taxon specific training, test and validation sets at a 60%, 20%, 20% split respectively as described in code at [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0008_split_and_encode.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0008_split_and_encode.Rmd)

@@

To visualise the regions of the sequences that the models were using to discriminate between effector and non-effector we used a GRAD-CAM approach ^[Ruth, there is absolutely no description of this in your report, please add a description of what you did, with which inputs and settings and versions etc].

@@

We selected four base architectures on which to build models for learning. Two of these contained Convolutional Neural Network (CNN) layers followed by either a Long Short Term Memory Layer (LSTM) or a Gated Recurrent Unit (GRU), two contained an Embedding Layer followed by the LSTM or GRU. All models had fully-connected dense layers after this. See Figure \@ref(fig:modelarch). Hyperparameter values for the models were identified using a random search as described in methods. Model variants within the hyperparameter search were assessed by comparing accuracy values on the development validation fraction of the training data. For each model type and taxon training data combination we selected the hyperparameter set giving highest accuracy on the validation set for manual tuning.  Final selected hyperparameter settings for models in each taxon can be seen in Table \@ref(tab:hyperparam), ^[Ruth, what were the actual best accuracy and loss values for the hyperparamter scan models selected and in Table \@ref(tab:hyperparam) , this table should have a column that shows this] a total of twelve models were produced, one for each taxon and model combination.

We then manually ran and checked the accuracy and loss of the models on the training and validation sets to investigate instances of overfitting and assess generality. Smaller models are less likely to overfit data, so we investigated the effect of regularization rate, filter count and kernel size on the progress and accuracy of the model as we reduced the size. We found that by reducing the number of kernels on all models from 2 to 1 and the number of filters reduced from 32 to 16 we removed apparent overfitting and retained high accuracy, with training completing in 40 epochs. Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress). We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4 % to 97.4 %, for the four fungal models we observed accuracy in the range 60.5 % to 84.2 % and for the four oomycete models we saw accuracy from 64.7 % to 82.3 %, reported in Fig \@ref(fig:overallmodel). All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.

@@

We examined the tendency of the models to call false positives or false negatives preferentially by creating confusion matrices of the classifications relative to the ground truth on the hold-out test data.\@ref(fig:confusionfinalmodels) The bacterial sequence trained models in general showed high accuracy and only one or two miscalls with no error bias except for the GRU-Embedding model which called five from 38 effectors as non effectors. The fungal sequence trained models were less accurate overall and showed a small amount more bias, again in the GRU-Embedding model, which was biased towards calling effectors as non-effectors and the CNN-LSTM model which was slightly biased in the opposite direction, calling non-effectors as effectors. The oomycete models were again quite balanced but the GRU-Embedding model showed a quite conservative tendancy calling 12 out of 17 effectors as non-effectors whilst getting all 17 non-effectors correct. Overall the models are accurate and show little to no bias toward false positive or false negatives, with the exception of the GRU-Embedding type. In oomycete sequences in particular and in this class of model across the different sequence types showed itself to tend to call real effectors as not of that class.

Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range 0.8 - 0.88, see Figure \@ref(fig:modelcorrs). CNN-GRU and LSTM-Embedding showed identical prediction sets. We saw similar levels of correlation  in the CNN-LSTM, GRU-Embedding and CNN-GRU fungal sequence trained model, in the range 0.79 - 0.9; though there was a significantly lower range of correlations with the LSTM-Embedding which were in the range 0.36 - 0.51. The models trained on oomycete sequences all showed this lower range of correlations, in the range 0.3 - 0.65. The higher correlation across bacterial trained models is likely from a mixture of the larger training set size and a greater uniformity of the sequences themselves. For the fungal sequence trained models we can see that the LSTM-Embedding model does not perform as well as the others. The oomycete sequence trained models all show a lower range correlation reflecting the likely less uniform and smaller training set. It is clear that, particularly for the fungal and oomycete models, each architecture is capturing separate aspects of the sequences and classifying on those with slightly varying levels of success.

@@

Given the accuracies of the above we selected the ensemble bacterial model and the CNN-LSTM fungal and oomycete models ^[Ruth, can you confirm that this is accurate for the rest of the analysis - are these the combinations of models you are using: You will need to discuss this with me before we make any changes to this document] to evaluate the performance of our models against widely used effector identification software. We compared against predictions from the bacterial effector prediction programs deepT3 [@deept3] and effectiveT3 [@effectivet3], the fungal effector prediction programs EffectorP 1 and 2 [@effectorp, @effectorp2] and the oomycete effector predictor effectR [@effectR]. Each comparison was carried out using the respective hold-out test sequence set for each taxon. For all taxon we saw greater Accuracy and F1 scores from our models than the established software, as shown in Figure \@ref(fig:vsothersmetrics). This was particularly marked in the F1 score, which incorporates a measure of the incorrect calls. Absoloute improvements were up to ~ 18 % in bacterial sequences, ~15 % in fungal sequences and ~ 28% in the oomycete sequences. The confusion matrices in Figure \@ref(fig:vsotherscfmat) show that accuracy and F1 score was compromised in all the established tools by the tendency of them all to misclassify true effectors as not effectors.  All the established software classifiers we tested show lower sensitivity than the models we have developed here.

^[Ruth, these confusion matrices have different axis orders to the ones earlier. Can you make them consistent?]

@@

An advantage of CNNs relative to other deep neural networks is their relative interpretability. A CNN can be analysed to activation maps which  highlight the regions of the input to which the CNN is most strongly using to classify. To examine the responses of the model, we ran the effector sequence ^[Ruth, can you confirm that this is what you did, we want the effector activations only, not the whole set, or at least the effector activations in one panel and the non-effectors in another] data sets back through the CNN-LSTM models for each taxon and extracted the network activations using the GRAD-CAM method. The profiles were smoothed using FFT and examined \@ref(fig:activations). All the models showed a peak of activation at the N-terminus of the sequences, coincident with expected positions of secretion signals. The fungal sequences created a single broad activation region with a width of 50 to 100 amino acids while the bacterial and oomycete sequences create a some smaller grouped peaks in a broader region which were each around 20 amino acids. Compositional analysis of the sequence under the peaks showed no apparent primary sequence conservation or motifs even within a taxon and data set. The sequences in the activation regions in all the taxa and models we examined did seem to be preferentially made up of hydrophobic and polar amino acids, which is apparently contradictory

@@

To make the models useful for developers in their own analytic pipelines we have provided an R package that provides a useful interface to the models. The package and installation instructions are available from GitHub [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff)

```{r, effsummary, echo=FALSE,message = FALSE}

apa_table(
  readr::read_tsv("tables/effectors_by_path_group.tab"),
  digits = 0,
  align = 'c',
  caption = 'Count of effectors listed in publications curated by PHI-Base used in this study in three major plant pathogen groups'
)
```

```{r, modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C:LSTM-Embedding model; D:GRU-Embedding model", message = FALSE, echo=FALSE}
knitr::include_graphics("figures/model_arch.png")
```

```{r, trainingprogress, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set", message = FALSE, echo=FALSE}
knitr::include_graphics("figures/final_train_progress.png")
```

```{r, overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.", echo=FALSE, message=FALSE}

# apa_table(
#   readr::read_tsv("tables/final_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.'
# )
readr::read_tsv("tables/final_model_performance_tidy.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Model, Accuracy) +
  ggplot2::geom_point(ggplot2::aes(colour = Accuracy)) +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x") 
```

```{r confusionmodelcorrs, fig.height = 8, fig.width = 8, fig.cap = "Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data; and Pearson correlations between classifications made by different models. A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.", message = FALSE, echo=FALSE}
# Load data
bacteria_ensemble_results <- data.table::fread("data/raw/bacteria_pred_ens.csv", drop = "V1")

bacteria_test_label <- data.table::fread("data/raw/bacteria_testing.csv") %>%
  dplyr::select(label) %>%
  mutate_each(list(as.factor))

@@

oomycete_test_label <- data.table::fread("data/raw/oomycete_testing.csv") %>%
  dplyr::select(label) %>%
  mutate_each(list(as.factor))

@@

fungi_test_label <- data.table::fread("data/raw/fungi_testing.csv") %>%
  dplyr::select(label) %>%
  mutate_each(list(as.factor))

@@

# Confusion matrices
conf_bacteria <- plot_confusion_matrices(
  data = bacteria_ensemble_results,
  true_label = bacteria_test_label,
  model_list = c("cnn_lstm", "cnn_gru", "gru_emb", "lstm_emb"),#, "ensemble_weighted", "ensemble_voting")
  fill_upper_limit = NA
)

conf_fungi <- plot_confusion_matrices(
  data = fungi_ensemble_results,
  true_label = fungi_test_label,
  model_list = c("cnn_lstm", "cnn_gru", "gru_emb", "lstm_emb"),#, "ensemble_weighted", "ensemble_voting")
  fill_upper_limit = NA
)

conf_oomycete <- plot_confusion_matrices(
  data = oomycete_ensemble_results,
  true_label = oomycete_test_label,
  model_list = c("cnn_lstm", "cnn_gru", "gru_emb", "lstm_emb"),#, "ensemble_weighted", "ensemble_voting")
  fill_upper_limit = NA
)

# Patchwork
corr_patch <- (
  corr_bacteria + conf_bacteria +
    plot_layout(ncol = 2, widths = c(1,3), tag_level = "new")
  ) /
  (
    corr_fungi + conf_fungi +
      plot_layout(ncol = 2, widths = c(1,3), tag_level = "new")
  ) /
  (
    corr_oomycete + conf_oomycete +
      plot_layout(ncol = 2, widths = c(1,3), tag_level = "new")
  )

corr_patch +
  plot_annotation(tag_levels = c("A", "1"), tag_sep = ".")
```

```{r ensembleresults, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.", message = FALSE, echo=FALSE}

# apa_table(
#   readr::read_tsv("tables/ensemble_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.'
readr::read_tsv("tables/ensemble_model_performance_tidy.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Combination, Accuracy) +
  ggplot2::geom_point(ggplot2::aes(colour = EnsembleFunction)) +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x")
```

```{r vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software", message = FALSE, echo=FALSE}
readr::read_tsv("tables/cfothertools.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Classifier, Score) +
  ggplot2::geom_point(ggplot2::aes(colour = Metric)) +
  ggplot2::facet_wrap(~ Group, scales = "free_x")
```

```{r vsotherscfmat, fig.width=9, fig.height=8, fig.cap="Confusion matrices of classifications made by our models compared with those made by widely used software. A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences.", message = FALSE, echo=FALSE}
# knitr::include_graphics("figures/cf_other_tools.png")

>>>>

---
title: "Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    # template: svm_latex.tex
    toc: no
fontsize: 12pt
fontfamily: sans
papersize: a4paper
author:
  - name: Ruth Kristianingsih, Dan MacLean
  - affiliation: The Sainsbury Laboratory, University of East Anglia, Norwich Research Park, Norwich, UK, NR4 7JJ, dan.maclean@tsl.ac.uk
abstract: "TBD."
geometry: margin=1in
biblio-style: apsr
bibliography: ["references.bib", "r-packages.bib"]
graphics: yes
header-includes:
  - \usepackage{amsmath}
  - \usepackage{siunitx}
---

@@

# Source plotting functions
source("functions/plot_confusion_mat.R")
source("functions/plot_confusion_matrices.R")
source("functions/plot_acc_loss_history.R")
source("functions/get_lower_corr_mat.R")
source("functions/get_seqlogo.R")

# Build R pacakges Bib build
r_refs(file = "r-packages.bib")

@@

```{r analysis-preferences, echo=FALSE}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(
  cache.extra = knitr::rand_seed,
  message = FALSE,
  echo = FALSE
)
```

@@

Sequence data were collected from the PHI-Base database version 4.8 [@phibase] by accessing a text dump of the data kindly prepared on request by the PHI-Base team, the file can be accessed at [https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv). The pipeline in Figure \@ref(fig:effectorcollection) outlines the steps used. We filtered plant effector proteins and their taxonomic groups and collected sequences from UniProt Release 2019_05, using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd). We created a correspondingly sized data set of non-effectors with secretion signals originating in species matched to those from which the effectors were drawn. We downloaded sequences for randomly selected proteins matching these criteria from Ensembl databases [@yates2020] : specifically Ensembl Fungi, Protists and Bacteria manually using the BioMart tools [@smedleybiomart2009]. Since the BioMart tool is not available on Ensembl Bacteria, we downloaded whole proteome protein sequnces from species matched to those from which the effector came using FTP. With these we used SignalP 3.0 [@dyrlovbendtsen2004] in order to filter the secreted sequences and selected accordingly. We used default paramaters from SignalP, except the type of organism group which is `euk` for both fungi and oomycete sequences, and `gram-` or `gram+` for bacteria sequences. Redundant sequences were filtered using BLASTp [@camachoblast2009]. We achieved these steps using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd).

```{r effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
knitr::include_graphics(
  "figures/flowchart_effectorsequence_collection.pdf"
)
```

```{r noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists"}
knitr::include_graphics(
  "figures/flowchart_secreted_noneffectorsequence_collection.pdf"
)
```

@@

The sequences collected were encoded using either one-hot encoding (CNN-LSTM based models) or integer based encoding (CNN-GRU-LSTM models). Sequences were post-padded with zeroes to bring the vectors to identical lengths to each other and the longest sequence in the taxon data set. The longest sequence  for bacteria, fungi, and oomycete are 2574, 4034, and 934, respectively. Encoded sequences were split into taxon specific training, test and validation sets at a 60%, 20%, 20% split respectively as described in code at [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0008_split_and_encode.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0008_split_and_encode.Rmd)

@@

To visualise the regions of the sequences that the models were using to discriminate between effector and non-effector we adapted a Grad-CAM (Gradient-weighted Class Activation Mapping) approach [@Selvaraju2019]. Using this method, we could produce the visualization of one of the convolutional neural network output values when it does a prediction. This will allow us to see which position in peptide sequences is important in determining whether it is effector or non-effector.

To find the importance of sequence position detected by our model, we took the output feature map of a convolution layer, given a sequence input. The convolutional layer we chose is the `convd_1` layer. Then, we computed the gradient of model output with the respect of the output of convolutional layer we chose. After that, we weighed every channel of that feature map with the gradient we calculated earlier, and took an average of the result.

Following the process mentioned above, we then summed all of the activation map results of all sequences in our datasets for each taxon and normalised them. Next, the final summed activation map for each taxon was smoothed using fourier analysis and plotted the smoothed version for each of them. The code on computing and visualising these heatmaps can be found on [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/heatmaps](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/heatmaps) .

@@

We selected four base architectures on which to build models for learning. Two of these contained Convolutional Neural Network (CNN) layers followed by either a Long Short Term Memory Layer (LSTM) or a Gated Recurrent Unit (GRU), two contained an Embedding Layer followed by the LSTM or GRU. All models had fully-connected dense layers after this. See Figure \@ref(fig:modelarch). Hyperparameter values for the models were identified using a random search as described in methods. Model variants within the hyperparameter search were assessed by comparing accuracy values on the development validation fraction of the training data. For each model type and taxon training data combination we selected the hyperparameter set giving highest accuracy on the validation set for manual tuning.  Final selected hyperparameter settings for models in each taxon can be seen in Table \@ref(tab:hyperparam), a total of twelve models were produced, one for each taxon and model combination. The values of accuracy and loss of each model produced are shown in Table \@ref(tab:acclossfinal).

We then manually ran and checked the accuracy and loss of the models on the training and validation sets to investigate instances of overfitting and assess generality. Smaller models are less likely to overfit data, so we investigated the effect of regularization rate, filter count and kernel size on the progress and accuracy of the model as we reduced the size. We found that by reducing the number of kernels on all models from 2 to 1 and the number of filters reduced from 32 to 16 we removed apparent overfitting and retained high accuracy, with training completing in 40 epochs. Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress). We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range $\sim 93.4 \%$  to  $\sim 97.4 \%$, for the four fungal models we observed accuracy in the range $\sim 60.5 \%$  to  $\sim 84.2 \%$  and for the four oomycete models we saw accuracy from $\sim 64.7 \%$  to  $\sim 82.3 \%$, reported in Fig \@ref(fig:overallmodel). All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.

@@

We examined the tendency of the models to call false positives or false negatives preferentially by creating confusion matrices of the classifications relative to the ground truth on the hold-out test data. The bacterial sequence trained models in general showed high accuracy and only one or two miscalls with no error bias except for the GRU-Embedding model which called five from 38 effectors as non effectors. The fungal sequence trained models were less accurate overall and showed a small amount more bias, again in the GRU-Embedding model, which was biased towards calling effectors as non-effectors and the CNN-LSTM model which was slightly biased in the opposite direction, calling non-effectors as effectors. The oomycete models were again quite balanced but the GRU-Embedding model showed a quite conservative tendancy calling 12 out of 17 effectors as non-effectors whilst getting all 17 non-effectors correct. Overall the models are accurate and show little to no bias toward false positive or false negatives, with the exception of the GRU-Embedding type. In oomycete sequences in particular and in this class of model across the different sequence types showed itself to tend to call real effectors as not of that class.

Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range \numrange[range-phrase = --]{0.80}{0.88}, see Figure \@ref(fig:confusionmodelcorrs). CNN-GRU and LSTM-Embedding showed identical prediction sets. We saw similar levels of correlation  in the CNN-LSTM, GRU-Embedding and CNN-GRU fungal sequence trained model, in the range \numrange[range-phrase = --]{0.79}{0.90}; though there was a significantly lower range of correlations with the LSTM-Embedding which were in the range \numrange[range-phrase = --]{0.36}{0.51}. The models trained on oomycete sequences all showed this lower range of correlations, in the range \numrange[range-phrase = --]{0.30}{0.65}. The higher correlation across bacterial trained models is likely from a mixture of the larger training set size and a greater uniformity of the sequences themselves. For the fungal sequence trained models we can see that the LSTM-Embedding model does not perform as well as the others. The oomycete sequence trained models all show a lower range correlation reflecting the likely less uniform and smaller training set. It is clear that, particularly for the fungal and oomycete models, each architecture is capturing separate aspects of the sequences and classifying on those with slightly varying levels of success.

@@

Given the accuracies of the above we selected the ensemble bacterial model and the CNN-LSTM fungal and oomycete models to evaluate the performance of our models against widely used effector identification software. We compared against predictions from the bacterial effector prediction programs deepT3 [@deept3] and effectiveT3 [@effectivet3], the fungal effector prediction programs EffectorP 1 and 2 [@effectorp, @effectorp2] and the oomycete effector predictor effectR [@effectR]. Each comparison was carried out using the respective hold-out test sequence set for each taxon. For all taxon we saw greater Accuracy and F1 scores from our models than the established software, as shown in Figure \@ref(fig:vsothersmetrics). This was particularly marked in the F1 score, which incorporates a measure of the incorrect calls. Absoloute improvements were up to $\sim 15\%$  in bacterial sequences, $\sim 15\%$ in fungal sequences and $\sim 20\%$  in the oomycete sequences. The confusion matrices in Figure \@ref(fig:vsotherscfmat) show that accuracy and F1 score was compromised in all the established tools by the tendency of them all to misclassify true effectors as not effectors.  All the established software classifiers we tested show lower sensitivity than the models we have developed here.

@@

An advantage of CNNs relative to other deep neural networks is their relative interpretability. A CNN can be analysed to activation maps which  highlight the regions of the input to which the CNN is most strongly using to classify. To examine the responses of the model, we ran the effector sequence data sets back through the CNN-LSTM models for each taxon and extracted the network activations using the Grad-CAM method. The profiles were smoothed using FFT and examined \@ref(fig:activations). All the models showed a peak of activation at the N-terminus of the sequences, coincident with expected positions of secretion signals. The fungal sequences created a single broad activation region with a width of 50 to 100 amino acids while the bacterial and oomycete sequences create a some smaller grouped peaks in a broader region which were each around 20 amino acids. Compositional analysis of the sequence under the peaks showed no apparent primary sequence conservation or motifs even within a taxon and data set. The sequences in the activation regions in all the taxa and models we examined did seem to be preferentially made up of hydrophobic and polar amino acids, which is apparently contradictory

@@

To make the models useful for developers in their own analytic pipelines we have provided an R package that provides a useful interface to the models. The package ans installation instructions are available from GitHub [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff).

```{r, effsummary}
apa_table(
  readr::read_tsv("tables/effectors_by_path_group.tab"),
  digits = 0,
  align = 'c',
  caption = 'Count of effectors listed in publications curated by PHI-Base used in this study in three major plant pathogen groups'
)
```

```{r modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C: LSTM-Embedding model; D: GRU-Embedding model"}
knitr::include_graphics("figures/model_arch.pdf")
```

```{r trainingprogress, fig.height= 10, fig.width= 10, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set"}
# knitr::include_graphics("figures/final_train_progress.png")

bacteria_acc_loss <- plot_acc_loss_history(
  train_data = data.table::fread("data/processed/bacteria_hist_train.csv"),
  val_data = data.table::fread("data/processed/bacteria_hist_val.csv"),
  plot_tag = "A"
) + ggplot2::theme(legend.position = "none", axis.title.x=element_blank())

fungi_acc_loss <- plot_acc_loss_history(
  train_data = data.table::fread("data/processed/fungi_hist_train.csv"),
  val_data = data.table::fread("data/processed/fungi_hist_val.csv"),
  plot_tag = "B"
) +
  ggplot2::theme(strip.background.x = element_blank(),
   strip.text.x = element_blank(),
    axis.title.x = element_blank())

oomycete_acc_loss <- plot_acc_loss_history(
  train_data = data.table::fread("data/processed/oomycete_hist_train.csv"),
  val_data = data.table::fread("data/processed/oomycete_hist_val.csv"),
  plot_tag = "C"
) +
  ggplot2::theme(legend.position = "none") +
  ggplot2::theme(strip.background.x = element_blank(),
   strip.text.x = element_blank())

(bacteria_acc_loss + fungi_acc_loss + oomycete_acc_loss) +
  plot_layout(nrow = 3, heights = c(3, 3, 3))
```

```{r overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences."}
# apa_table(
#   readr::read_tsv("tables/final_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.'
# )
readr::read_tsv("tables/final_model_performance_tidy.tab") %>%
  ggplot2::ggplot() +
  ggplot2::aes(
    x = Model,
    y = Accuracy,
    color = Accuracy
  ) +
  ggplot2::geom_point() +
  ggplot2::scale_color_viridis_c(begin = 0.65, end = 0.1) +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x") +
  ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r confusionmodelcorrs, fig.height = 8, fig.width = 8, fig.cap = "Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data; and Pearson correlations between classifications made by different models. A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
# Load data
bacteria_ensemble_results <- data.table::fread("data/raw/bacteria_pred_ens.csv", drop = "V1")

bacteria_test_label <- data.table::fread("data/raw/bacteria_testing.csv") %>%
  dplyr::select(label) %>%
  dplyr::mutate(label = as.factor(label))

@@

oomycete_test_label <- data.table::fread("data/raw/oomycete_testing.csv") %>%
  dplyr::select(label)  %>%
  dplyr::mutate(label = as.factor(label))

@@

fungi_test_label <- data.table::fread("data/raw/fungi_testing.csv") %>%
  dplyr::select(label)  %>%
  dplyr::mutate(label = as.factor(label))

@@

# Confusion matrices
conf_bacteria <- plot_confusion_matrices(
  data = bacteria_ensemble_results,
  true_label = bacteria_test_label,
  model_list = c("cnn_lstm", "cnn_gru", "gru_emb", "lstm_emb"),#, "ensemble_weighted", "ensemble_voting")
  fill_upper_limit = NA
) +
  ggplot2::labs(tag = "A")

conf_fungi <- plot_confusion_matrices(
  data = fungi_ensemble_results,
  true_label = fungi_test_label,
  model_list = c("cnn_lstm", "cnn_gru", "gru_emb", "lstm_emb"),#, "ensemble_weighted", "ensemble_voting")
  fill_upper_limit = NA
) +
  ggplot2::labs(tag = "B")

conf_oomycete <- plot_confusion_matrices(
  data = oomycete_ensemble_results,
  true_label = oomycete_test_label,
  model_list = c("cnn_lstm", "cnn_gru", "gru_emb", "lstm_emb"),#, "ensemble_weighted", "ensemble_voting")
  fill_upper_limit = NA
) +
  ggplot2::labs(tag = "C")

# Patchwork
corr_patch <- (
  conf_bacteria + corr_bacteria +
    plot_layout(ncol = 2, widths = c(3,1), tag_level = "new")
  ) /
  (
    conf_fungi +  corr_fungi +
      plot_layout(ncol = 2, widths = c(3,1), tag_level = "new")
  ) /
  (
    conf_oomycete + corr_oomycete +
      plot_layout(ncol = 2, widths = c(3,1), tag_level = "new")
  )

corr_patch +
  plot_annotation(tag_levels = c("A"))
```

```{r ensembleresults, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences."}
# apa_table(
#   readr::read_tsv("tables/ensemble_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.'
readr::read_tsv("tables/ensemble_model_performance_tidy.tab") %>%
  ggplot2::ggplot() +
  ggplot2::aes(x = Combination, y = Accuracy) +
  ggplot2::geom_point(ggplot2::aes(colour = EnsembleFunction,
                                   shape = EnsembleFunction), size = 2) +
  ggplot2::scale_shape_manual(values = c("Ensemble Voting" = 19, "Ensemble Weighted" = 17)) +
  ggplot2::scale_color_viridis_d(begin = 0.65, end = 0.1) +
  ggplot2::labs(color = "Ensemble Function", shape = "Ensemble Function") +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```
